---
title: 'Project 1: Wrangling, Exploration, Visualization'
author: "SDS322E"
date: ''
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
  pdf_document:
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", warning = F, message = F,
tidy=TRUE, tidy.opts=list(width.cutoff=60), R.options=list(max.print=100))
```

## Data Wrangling, Exploration, Visualization

### CJL3526 Carson Lansdowne

## Analysis of Satellite Data

This is a project (1) for SDS 322E, where the UCS Satellite Dataset and Space Track's public decay epoch calculation datasets are merged and analysed. 

The UCS Satellite Dataset is an exhaustive dataset listing the orbital parameters, launch info including country of origin, launch site, contractor, launch date, and launch vehicle, satellite purpose and other descriptions such as orbit class that is generated by a group of scientists volunteering their efforts. It is regarded as the primary (and sole) public satellite database and is added to regularly. This dataset is particularly useful because it contains the orbital parameters that are necessary to determine where in the sky a satellite is and how fast it is moving. Notably, it does not include the expected lifetime for all satellites it has listed -- this is included in the Space Track database, a commercial database provided online via an API that includes the expected orbital decay parameters for a given satellite, including historical satellites that have already de-orbited. Importantly, each satellite has a distinct NORAD (North American Aerospace Defense Command) identifier that is used to merge the datasets. These data are acquired by tedious scraping of the NORAD database, independent research & observations, and complex orbital mechanical simulations for the decay epoch timing.

The UCS Database is free to download and update online, and the Space Track database was acquired by setting up an account with the API service. I chose to study these databases because of my interest in orbital infrastructure and the space launch industry -- as an aerospace engineer, I am excited to enter the workforce and apply the data analytics savvy I am learning at University to this incredible industry. It will be shown in the data that satellites have been orbiting by the plethora since the late 20th century and I expect to see a significant increase in commercial satellites as well as commercial launches in the most recent two decades. It will also likely be apparent that there are clear relationships between larger satellites and higher orbits as well as the number of satellites in different classes of orbits. I also expect NASA owned launches to make up the majority of government launches across the board and explore this in the exploration section of the report. 

## Import Data

```{R}
library(tidyverse)
data1 <- read_csv("./Decay-Sat-Data.csv")
data2 <- read_csv("./UCS-Sat-Data.csv")

glimpse(data1)
glimpse(data2)
```
## Initial assessment 
```{R}
library(dplyr)
data1_row <- nrow(data1)
data2_row <- nrow(data2)
data1_distinct <- n_distinct(data1$'NORAD Number')
data2_distinct <- n_distinct(data2$'NORAD Number')
common_NORAD <- data2 %>% inner_join(data1, by="NORAD Number")
common_row <- n_distinct(common_NORAD$`NORAD Number`)
unique_data1 <- data1 %>% anti_join(data2, by="NORAD Number")
unique_data1_row <- n_distinct(unique_data1$`NORAD Number`)
unique_data2 <- data2 %>% anti_join(data1, by="NORAD Number")
unique_data2_row <- n_distinct(unique_data2$`NORAD Number`)

```

## Tidying and Merging

Because the datasets were already tidy, the Space Track dataset was pivoted wider into columns for each country and using the values from the decay epoch timing. This creates a large dataset with a column for each country and mostly NULL values in the values for decay epoch -- each satellite in the row only has one country of origin/maintenance. When pivoted longer, the country column is populated with these column names and the values are pulled into the decay epoch. The NULL and na values for the decay epoch column must then be cleaned from the dataset.

The joined dataset is 4759 observations compared to the 69525 observations of Space Track Data (data1) and the 4087 observations of the UCS dataset (data2). The left join is conducted on the UCS dataset with the Space Track data, and only a few (115) of the distinct NORAD IDs appear in both datasets, and the joined sat_data set is not much larger than the original UCS dataset -- the former is 787 observations larger because the Space Track data set has repeat NORAD numbers. UCS has 4078 unique NORAD numbers, Space Track has 25921, and sat_data has 4078. UCS has 3963 NORAD IDs not in the other and Space Track has 25806 not in UCS. These are displayed below.

```{R}

data2 <- data2[ -c(29, 38:255)]
data1 <- data1 %>% pivot_wider(names_from='COUNTRY', values_from='DECAY_EPOCH')
data1 <- data1 %>% pivot_longer(12:73, names_to='COUNTRY', values_to='DECAY_EPOCH')
data1 <- data1[!(is.na(data1$DECAY_EPOCH) | data1$DECAY_EPOCH=="NULL"), ]

glimpse(data1)

sat_data <- data2 %>% left_join(data1, by="NORAD Number")

glimpse(sat_data)

#common IDs
glimpse(common_NORAD)

#data 1 unique IDS
glimpse(unique_data1)

#data 2 unique IDS
glimpse(unique_data2)

```
## Data Exploration

The data is explored and rather interesting results are found here -- Many of the satellites are expected to last less than a year but the longest lived of the 4,000+ are expected to last over 30 years. The majority of NASA launches span over a decade, though in the middle part of the past 20 years the mean NASA launches were expected to last only less than 5 years. It is likely that the organization was focusing on experimental launches during this time. The Mass/year term is found in the sat_data set as a metric attempting to "weight" the mass of each satellite by normalizing it for the amount of years it is in orbit around the planet. Mass is extremely difficult to get into outer space and so the satellites that have very high Mass per Year can be considered to be much more valuable, or worth the trouble of getting into orbit. According to this metric, the most "valuable" satellite is 1HOPSAT, a National Reconnaissance satellite launched by the USA in 2021.
```{R}
#exploration of data

lifetime_sort <- sat_data %>% group_by(`Class of Orbit`) %>% filter(`Country of Operator/Owner` == 'USA') %>% arrange(`Expected Lifetime (yrs.)`)
decay_clearing <- sat_data %>% filter(!str_detect(DECAY_EPOCH, "NA")) %>% filter(!str_detect(`Expected Lifetime (yrs.)`, "NA")) %>% arrange(`Date of Launch`)
sat_data <- sat_data %>% mutate(Date = `Date of Launch`) %>% separate('Date',into=c("Month","Day","Year"),sep="/")
NASA_launches <- sat_data %>% select(`Launch Mass (kg.)`,`Expected Lifetime (yrs.)`,`Year`,`Operator/Owner`) %>% filter(str_detect(`Operator/Owner`,".NASA.")) %>% group_by(`Year`) %>% summarize_all(mean,na.rm=T)
sat_data <- sat_data %>% mutate('Mass per Year' = `Launch Mass (kg.)`/`Expected Lifetime (yrs.)`) 
```

## Summary Statistics

The procedure for finding the summary statistics is as follows:

The numerical values of the sat_data set are isolated and analyzed using summarize_all for mean, min, max, sd, and IQR. Additionally, the proportion of values greater than the mean is found using a custom function. These data are displayed on the table below -- notably, the sd is relatively high compared to the mean of most of the numerical data, suggesting that the orbital characteristics of all of the satellites humanity has in orbit are quite varied. The only metric where this is not the case is launch mass.

Additionally, the number of na values in each column of the sat_data set is found and the count of Orbit types, orbit classes, and purposes of satellites is found and shown below. Finally, the mean and sd of satellite mass in kg when grouped by type of orbit and class of orbit is found and shown below. Type of orbit seems to have a relatively significatn effect on the statistical values of the satellites.

```{R}
#summary stats

sat_sum_mean <- sat_data %>% select(`Longitude of GEO (degrees)`,`Perigee (km)`,`Apogee (km)`,Eccentricity,`Inclination (degrees)`,`Period (minutes)`,`Launch Mass (kg.)`,`Expected Lifetime (yrs.)`) %>% summarize_all(mean, na.rm=T) 
sat_sum_min <- sat_data %>% select(`Longitude of GEO (degrees)`,`Perigee (km)`,`Apogee (km)`,Eccentricity,`Inclination (degrees)`,`Period (minutes)`,`Launch Mass (kg.)`,`Expected Lifetime (yrs.)`) %>% summarize_all(min, na.rm=T)
sat_sum_max <- sat_data %>% select(`Longitude of GEO (degrees)`,`Perigee (km)`,`Apogee (km)`,Eccentricity,`Inclination (degrees)`,`Period (minutes)`,`Launch Mass (kg.)`,`Expected Lifetime (yrs.)`) %>% summarize_all(max, na.rm=T)
sat_sum_sd <- sat_data %>% select(`Longitude of GEO (degrees)`,`Perigee (km)`,`Apogee (km)`,Eccentricity,`Inclination (degrees)`,`Period (minutes)`,`Launch Mass (kg.)`,`Expected Lifetime (yrs.)`) %>% summarize_all(sd, na.rm=T)
sat_sum_IQR <- sat_data %>% select(`Longitude of GEO (degrees)`,`Perigee (km)`,`Apogee (km)`,Eccentricity,`Inclination (degrees)`,`Period (minutes)`,`Launch Mass (kg.)`,`Expected Lifetime (yrs.)`) %>% summarize_all(IQR, na.rm=T)


check_na <- function(x) {
  sum(is.na(x))
}

circularization <- function(x,y) {
  sum(x<2*y)
  
}

overmean <- function(x, na.rm=T){
  sum((x>mean(x,na.rm=T)),na.rm=T)/length(x) #proportion of values in the data over the mean
}

sat_sum_myfunc <- sat_data %>% select(`Longitude of GEO (degrees)`,`Perigee (km)`,`Apogee (km)`,Eccentricity,`Inclination (degrees)`,`Period (minutes)`,`Launch Mass (kg.)`,`Expected Lifetime (yrs.)`) %>% summarize_all(overmean)

sat_sum <- rbind(sat_sum_mean, sat_sum_min,sat_sum_max,sat_sum_sd,sat_sum_IQR,sat_sum_myfunc)


sat_sum_orbit_mean <- sat_data %>% group_by(`Type of Orbit`,`Class of Orbit`) %>% summarize(mean(`Launch Mass (kg.)`))
sat_sum_orbit_sd <- sat_data %>% group_by(`Type of Orbit`,`Class of Orbit`) %>% summarize(sd(`Launch Mass (kg.)`))

sat_sum_na <- sat_data %>% summarize_all(check_na)

sat_sum_orbit_type <- sat_data %>% group_by(`Type of Orbit`) %>% summarize(n())
sat_sum_orbit_class <- sat_data %>% group_by(`Class of Orbit`) %>% summarize(n())
sat_sum_purpose <- sat_data %>% group_by(Purpose) %>% summarize(n())

#install.packages("kableExtra")
library(kableExtra)

rownames(sat_sum) <- c("Mean","Min","Max","Std. Dev.","IQR","Over Mean %")
sat_sum %>% kbl(caption = "Summary Statistics of Satellite Data") %>% kable_minimal() 
```

Below are the total counts of "NAs" in each column within the satellite dataset as well as the count of several different groupings for noteable categorical variables.
```{R}
glimpse(sat_sum_na)
sat_sum_orbit_type
sat_sum_orbit_class
sat_sum_purpose
```

Below are the summary statistics mean and sd, respectively, for the satellite dataset numerical values sorted on Orbit type and class.
```{R}
sat_sum_orbit_mean
sat_sum_orbit_sd
```

## Visualization

The graphic below plots a histogram distribution over the past 45 years of the count of satellites that have been launched and are still in orbit for each of the users that UCS records and categorizes them in. The color fill is segmented by purpose of the machine in orbit -- it is immediately apparent that the count of satellites in orbit has skyrocketed in the pas decade, ramping heavily for military and government applications, but coming out of nowhere for civil in particular. Another trend to note is the increase in Earth observation efforts by the government in the early years of the 2010s and on. 

```{r fig.height=10, fig.width=12,error=TRUE}

library(ggplot2)

theme_set(theme_dark())

sat_data <- sat_data %>% group_by(Year,Purpose,Users) %>% mutate(count = n())

sat_data %>% ggplot(aes(x = Year, fill = Purpose))+geom_bar(alpha=.5)+geom_hline(yintercept = 0)+ facet_wrap(~`Users`,scales = "free_y", dir="v")+scale_x_discrete(breaks = scales::pretty_breaks(n = 15))+theme(legend.position = "bottom")+ggtitle("Satellite Purpose and launch count over time")+xlab("Epoch")+ylab("Launch Count")+ theme(axis.text.x = element_text(size=5, angle = 45, vjust = 0.5, hjust=1))+geom_text(aes(y=count/2,label=count),check_overlap = TRUE)

```
The graphic below depicts the density distribution of satellites in orbit by launch mass for each of the 4 distinct classes of orbit: Low Earth (LEO), Elliptical, Geostationary (GEO), and Medium Earth (MEO). The total distribution of the satellites is also shown with the blue density line. It is interesting that each of the classes seems to have it's own small domain of satellite masses, with the larger masses taking up the higher orbits. Interestingly, LEO class satellites do seem to span into significantly higher masses than Elliptical or MEO -- likely experimental satellites and ease of use case in LEO driving this.

```{R,error=TRUE}
theme_set(theme_minimal())

sat_data %>% ggplot(aes(x=`Launch Mass (kg.)`))+geom_density(alpha=.75,aes(fill=`Class of Orbit`))+theme(legend.position = "bottom")+geom_density(color="blue",alpha=.1)+ggtitle("Launch Mass (log scale) Density Distribution by Orbit Class")+geom_rug()+scale_x_continuous(n.breaks=10)

```
The graphic below depicts a zoomed in section of the 2d density mapping of apogee vs perigee for orbits. Perigee represents the lowest point on the orbit and Apogee the highest. When the values are equal, the orbit is considered circular. It is seen from the graph below that there is a focused distribution of mostly circular orbits but it is very small and narrow. The majority of orbits have highly elliptic paths and are homogenously distributed, for the most part, across the plane of Perigee/Apogee coordinates. Perigee is shown on a log scale for simplicity of the zoom in on the two brigh points representing near-circularity. 

```{R,error=TRUE}
theme_set(theme_classic())

sat_data %>% ggplot(aes(`Perigee (km)`,`Apogee (km)`))+geom_density2d_filled()+theme(legend.position = "bottom")+ggtitle("Satellite Perigee vs Apogee 2d Density plot")+geom_rug()+scale_y_continuous(trans='log10')+scale_x_continuous(trans='log10')+xlim(500,600)+ylim(500,600)+xlab("Perigee log10(km)")
```
The graphic below brilliantly shows the orbital characteristics of period (time around the planet in minutes) for every year since 1985 for every launch vehicle on record in the UCS database. The plot shows the launch mass of the satellite by the size of the point on the plot and the points are colored by class of orbit. Importantly, each datapoint is a measure of the average launch mass for each year for each rocket -- all 159 of them. One of the more obvious trends in the graphic is that the rocket's which more frequent launches over the years tend to have larger launch masses. Perhaps less immediately obvious, the LEO launches are seem to be extremely flexible for all sizes of payloads and on many of the rockets -- it is easy to see the scatter of the blue dots across the entire graphic. 

```{r fig.height=30, fig.width=12,error=TRUE}
theme_set(theme_grey())

sat_data %>% ggplot(aes(Year,`Period (minutes)`))+geom_jitter(stat="summary",aes(col=`Class of Orbit`, size=`Launch Mass (kg.)`))+facet_wrap(~`Launch Vehicle`,scales = "free_y",ncol=4)+scale_x_discrete(breaks = scales::pretty_breaks(n = 4))+ggtitle("Period of multiple orbit classes over past 4 decades by launch vehicle")
```



